---
title: "Hurricane Track & Intensity Investigation"
author: "Henry Blassingame"
date: "02 OCTOBER 2020"
output: 
  html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(gdata)
library(ggplot2)
library(gstat)
library(lubridate)
library(readr)
library(maps)
library(maptools)
library(RColorBrewer)
library(gridExtra)
library(repr)
library(dplyr)
library(caret)
library(randomForest)
library(MLmetrics)
library(tidyr)
library(rgdal)
library(rgeos)
library(plotly)
library(DALEX)
library(auditor)

```


## Read and format input data
``` {r data loading and cleanup, echo=FALSE}
# Read the data file.

options(warn=0)

hurricane_data <- read.csv("atlantic_updated.csv", header = TRUE, colClasses = c("character", "character", "integer", "character", "character", "character", "integer", "numeric", "numeric", "integer", "integer", "integer", "integer", "integer", "integer", "integer",  "integer", "integer", "integer", "integer", "integer", "integer", "integer"))

# The date and time data is a mess. Let's reformat it into one field:

hurricane_data$Year <- as.numeric(substr(as.character(hurricane_data$Date), 1,4))
hurricane_data$Month <- substr(as.character(hurricane_data$Date), 5,6)
hurricane_data$Day <- substr(as.character(hurricane_data$Date), 7,8)
hurricane_data$ID <- as.character(hurricane_data$ID)
hurricane_data$dateString <- as.Date(paste(hurricane_data$Month, hurricane_data$Day, hurricane_data$Year, sep = "/"), "%m/%d/%Y")
hurricane_data$timestamp <- as.POSIXct(paste(hurricane_data$dateString, hurricane_data$Time), format = "%Y-%m-%d %H:%M")

coordinates<-(as.data.frame(hurricane_data$Longitude))
coordinates<-cbind(coordinates, hurricane_data$Latitude)
out<-subset(coordinates, hurricane_data$Longitude < -100) ##Find the outliers

# Exclude data before 1900
# hurricane_data.new <- hurricane_data[,Year >= 1900]
hurricane_data.new <- hurricane_data %>% filter(Year >= 1900)
hurricane_data.new$YYYYMM<-as.numeric(do.call(paste, c(hurricane_data.new[c("Year", "Month")], sep = ""))) 
# Exclude storms with only one record
hurricane_data.new<-hurricane_data.new[ave(rep(1, nrow(hurricane_data.new)), hurricane_data.new$ID, FUN=length)>1,] 

# Replace any NA values in Category with 0
hurricane_data.new$Category [is.na(hurricane_data.new$Category)]  <- 0

# define and remove the columns we don't need
cruft <- c("Date", "dateString", "Event", "Status",    "Low.Wind.NE", "Low.Wind.SE","Low.Wind.SW", "Low.Wind.NW", "Moderate.Wind.NE", "Moderate.Wind.SE","Moderate.Wind.SW", "Moderate.Wind.NW", "High.Wind.NE", "High.Wind.SE","High.Wind.SW", "High.Wind.NW",  "Day")


trimmed_data <- (hurricane_data.new[,!(names(hurricane_data.new) %in% cruft)])
trimmed_data$Category [is.na(trimmed_data$Category)]  <- 0
trimmed_data$storm_num  <- as.numeric(substr(trimmed_data$ID,3,4))

# create training and test data sets for later on
storm_partition <- createDataPartition(trimmed_data[,"ID"], times=1, p=.7, list=FALSE)
train_set <- trimmed_data[storm_partition,]
test_set <- trimmed_data[-storm_partition,]

```
##Create a map of the US Coastline and plot landfall points

```  {r data visualization, echo=FALSE}

usa <- map("state", fill = TRUE, plot=FALSE) 
IDs <- sapply(strsplit(usa$names, ":"), function(x) x[1])
usa <- map2SpatialPolygons(usa, IDs=IDs, proj4string=CRS("+proj=longlat +datum=WGS84"))

# FCreate the coastline
crs  <- CRS("+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-110 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m") 
usa.p<-spTransform(usa, crs)
usa.p<-gBuffer(usa.p, byid=TRUE, width=1) 
usa.u<- gUnaryUnion(usa.p)
usa.l<-as(usa.u,'SpatialLines') 
crs.new<-CRS(proj4string(usa)) 
usa.l<-spTransform(usa.l, crs.new)

hurricane_data.ID<-unique(trimmed_data$ID) # Identifying the numbers of hurricanes on the dataset
storm_landfall<-NULL                       # Initiate a NULL matrix where to store the results recursively 

options(warn=-1)                           # Removing the warning messages coming from the krige function
for (i in 1:length(hurricane_data.ID)) {

  id.storm<-hurricane_data.ID[i]
  storm.i<-trimmed_data[trimmed_data$ID==id.storm,]
  
  # Transform the Hurricane track into a line
  storm.c<- data.frame(x=storm.i$Longitude,y=storm.i$Latitude)
  na.omit(storm.c)
  coordinates(storm.c)<-~x+y 
  proj4string(storm.c) <- CRS(proj4string(usa))
  storm.l <- SpatialLines(list(Lines(list(Line(coordinates(storm.c))),"X")))
  proj4string(storm.l) <- CRS(proj4string(usa)) ## attributing the coordinate system of the usa map (WGS84) to the ike track
  
  # Finding the landfall point of the hurricane (intersection between the track and the US coastline)
  cross<-gIntersection(storm.l,usa.l)
  
  # If there is intersection, identify the wind speed, date (year and month) and State at landfall
  
  if (length(cross)>0)
  {
    storm.int<-data.frame(x=storm.i$Longitude,y=storm.i$Latitude,wmax=storm.i$Maximum.Wind,tim=storm.i$YYYYMM)
    colnames(storm.int)<-c("Longitude","Latitude","Max_Speed","YYYYMM")
    
    # Convert the event track info into a spatial data frame
    coordinates(storm.int) <- ~Longitude+Latitude
    proj4string(storm.int) <- CRS(proj4string(usa))
    
    # Compute the interpolated values of the wind speed at landfall
    landfall.k <- krige(Max_Speed~1, storm.int, cross, debug.level = 0)
    landfall.w<-landfall.k$var1.pred
    
    # Compute the interpolated values of the Year month at landfall
    landfall.kt <- krige(YYYYMM~1, storm.int, cross, debug.level = 0)
    landfall.t<-landfall.kt$var1.pred
    
    landfall.p<-cbind(cross@coords,landfall.w,landfall.t)
    colnames(landfall.p)[3:4]<-c("Max.Wind","Time")
    
    # IChoose the landfal for the storm with the strongest wind.
    lfmax.p<-landfall.p[which.max(landfall.p[,3]),]
    lfmax.p<-t(as.matrix(lfmax.p))
    lfmax.p[,c("Time")]<-round(lfmax.p[,c("Time")]) ## rounding the time in case the landfall point happens between 2 months
    
    # Identifying the state where landfall occurs
    lfmax.p1<-data.frame(x=lfmax.p[1], y=lfmax.p[2], wmax=lfmax.p[3])
    coordinates(lfmax.p1)<-~x+y
    proj4string(lfmax.p1) <- CRS(proj4string(usa))
    crs  <- CRS("+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-110 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m") 
    lfmax.p1_pla<-spTransform(lfmax.p1, crs)
    usa_pla<-spTransform(usa, crs)
    dist<-gDistance(lfmax.p1_pla,usa_pla, byid=TRUE)
    state.name<-rownames(dist)[which.min(dist)]
    
    # storm info to append
    st.info<-storm.i[1,c("ID","Name")]
    st.info<-cbind(st.info,state.name,lfmax.p)
    st.info$Year<-substr(st.info$Time,1,4)
    st.info$Month<-substr(st.info$Time,5,6)
    st.info<-st.info[,-which(colnames(st.info)=="Time")]
    
    # Append storm info
    storm_landfall<-rbind(storm_landfall,st.info)
  }
} 
options(warn=0)


# Add Hurricane category 
brk<-c(-Inf,74,95,110,129,156,+Inf)
storm_landfall$cat<-cut(storm_landfall$Max.Wind,breaks=brk,include.lowest=T,include.highest=T,  labels=c(0,1,2,3,4,5))

# Convert the category to numeric
storm_landfall$cat=as.numeric(levels(storm_landfall$cat))[storm_landfall$cat]
storm_landfall$Year=as.numeric(storm_landfall$Year)

storm_landfall<-storm_landfall[storm_landfall$Max.Wind >= 75, ]

# Visualise the resulting matrix with the landfall points
head(storm_landfall)
```

## Map Hurricane Landfall Points
```{r, message=F, warning=F, fig.width=10}

geography <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showland = TRUE,
  landcolor = toRGB("gray95"),
  subunitcolor = toRGB("gray85"),
  countrycolor = toRGB("gray85"),
  countrywidth = 0.5,
  subunitwidth = 0.5
)

storm_colors <- c("blue", "green", "yellow", "orange", "red")

storm_plot <- plot_geo(storm_landfall, lat = ~y, lon = ~x) %>%
  add_markers(
   text = ~paste(Year,"<br />",Name,"<br />","Cat:",cat),hoverinfo="text",
    color = ~storm_landfall$cat, size = ~storm_landfall$Max.Wind,
   colors = storm_colors,
    marker = list(colorbar = list(len = 0.2, title = "Landfall Category"))
  ) %>%
  layout(
    title = 'US Hurricane Landfall points', geo = geography
  )
  
storm_plot

```

## Boxplot of the maximum wind speed at landfall for selected states

```{r}


storm_landfall1<-subset(storm_landfall,state.name %in% c("florida","louisiana","texas","mississippi", "alabama", "south carolina", "north carolina"))

options(warn = -1) 
storm_box <- plot_ly(storm_landfall1, y = ~Max.Wind, color = ~state.name, type = "box") %>%
  layout(title='Max Wind speed at landfall', yaxis = list(title = 'Max Wind speed [mph]'))
storm_box

```

## Plot of storms by Category 
### This looked better in Tableau than I could have made it in R, so I am reusing this reom my last project.

[Plot of Hurricanes by Year and Category](https://public.tableau.com/views/HurricaneData_15971058287080/StormsbyCategoryYear?:language=en&:display_count=y&:origin=viz_share_link)


## Test of Random Forest vs. Linear Regression to predict number of storms per year
``` {r}
# test linear regression against random forest
lm_model <- lm(storm_num ~ Time+Category+Latitude+Longitude+Maximum.Wind+Year+Month+timestamp , data = train_set)

rf_model = randomForest(storm_num ~ Category+Latitude+Longitude+Maximum.Wind+Year+Month+timestamp, data = train_set, ntree = 5)

test_set$scores = predict(rf_model, newdata = test_set)

summary(test_set)

#get residuals
lm_exp <- DALEX::explain(lm_model, label = "lm", data = train_set, y = train_set$storm_num)
rf_exp <- DALEX::explain(rf_model, label = "rf", data = train_set, y = train_set$storm_num)

lm_mr <- model_residual(lm_exp)
rf_mr <- model_residual(rf_exp)

# Compare accuracy of Random Forest vs. Linear Regression. Random Forest is more accurate
plot(rf_mr, lm_mr, type = "prediction", abline = TRUE)


predictions <- rf_model %>% predict(test_set)
RMSE <-  RMSE(predictions, test_set$storm_num)
R2 <- R2(predictions, test_set$storm_num)

# Print Root Mean Square Error of the prediction
RMSE

# Print R-Squared
R2

```


